---
title: "Praxis-BGM R Tutorial"
author: "Qiran Jia"
date: "2026-04-20"
output:
  html_document: default
  pdf_document: default
---

## Overview

This tutorial shows how to use the updated Praxis-BGM R interface from the
single wrapper script `R/praxis_bgm_interface.R`. It reproduces an R-based
workflow with:

- simulated source data
- transferred priors for target-domain fitting
- a no-prior comparison fit
- Bayes-factor feature selection through `praxis_bgm_bf_selection()`

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)

library(reticulate)
library(MASS)
library(mclust)

praxis_python <- Sys.getenv("RETICULATE_PYTHON", unset = "")
if (nzchar(praxis_python)) {
  use_python(praxis_python, required = TRUE)
} else {
  use_condaenv("Praxis_env", required = TRUE)
}

source("R/praxis_bgm_interface.R")
```

If needed, install the R packages used by this tutorial:

```{r install-packages, eval=FALSE}
install.packages(c("reticulate", "MASS", "mclust", "proxy", "clue", "rmarkdown"))
```

## Simulate source data

```{r source-sim}
K <- 4L
P <- 100L
C <- 40L
N_src <- 800L

source_data <- generate_overlapping_gmm_samples(
  n_components = K,
  n_causal = C,
  n_features = P,
  n_samples = N_src,
  random_seed = 123
)

X_src <- source_data$samples
y_src <- source_data$labels
true_means_src <- source_data$full_means

cat("X_src shape:", dim(X_src), "\n")
cat("True means (source) shape:", dim(true_means_src), "\n")
```

## Learn source priors

```{r source-priors}
prior_mus_src <- do.call(
  rbind,
  lapply(seq_len(K), function(k) {
    Xk <- X_src[y_src == (k - 1L), , drop = FALSE]
    if (nrow(Xk) == 0L) {
      stop(sprintf("No samples found for source cluster %d.", k - 1L))
    }
    colMeans(Xk)
  })
)

prior_weights_src <- as.numeric(table(factor(y_src, levels = 0:(K - 1L)))) / length(y_src)
prior_Sigmas_src <- source_data$covs

cat("Empirical source prior weights:", round(prior_weights_src, 3), "\n")
cat("Prior means shape:", dim(prior_mus_src), "\n")
```

## Supervised QDA baseline from source

```{r qda-source}
qda <- MASS::qda(X_src, grouping = as.factor(y_src))
cat("QDA fit complete.\n")
```

## Simulate target data with domain shift

```{r target-sim}
N_tgt <- 200L
mean_shift_tgt <- 0.3
shift_percentage <- 0.3

true_means_tgt <- randomly_shift_means(
  true_means_src,
  shift_magnitude = mean_shift_tgt,
  percentage = shift_percentage,
  random_seed = 999
)

set.seed(999)
target_sizes <- rep(N_tgt %/% K, K)
if (N_tgt %% K > 0L) {
  target_sizes[seq_len(N_tgt %% K)] <- target_sizes[seq_len(N_tgt %% K)] + 1L
}

X_tgt_list <- vector("list", K)
y_tgt_list <- vector("list", K)

for (k in seq_len(K)) {
  n_k <- target_sizes[k]

  causal_tgt <- MASS::mvrnorm(
    n_k,
    true_means_tgt[k, seq_len(C)],
    diag(C)
  )
  causal_tgt <- matrix(causal_tgt, nrow = n_k, ncol = C)

  non_causal_tgt <- matrix(
    rnorm(n_k * (P - C)),
    ncol = P - C
  )

  X_tgt_list[[k]] <- cbind(causal_tgt, non_causal_tgt)
  y_tgt_list[[k]] <- rep(k - 1L, n_k)
}

X_tgt <- do.call(rbind, X_tgt_list)
y_tgt <- unlist(y_tgt_list, use.names = FALSE)

cat("X_tgt shape:", dim(X_tgt), "\n")
cat("True means (target) shape:", dim(true_means_tgt), "\n")
```

## QDA baseline on target

```{r qda-target}
qda_preds <- predict(qda, X_tgt)$class
ari_qda <- mclust::adjustedRandIndex(y_tgt, as.integer(as.character(qda_preds)))

cat(sprintf("Target ARI for source-trained QDA: %.3f\n", ari_qda))
```

## Praxis-BGM with transferred priors

```{r praxis-prior-fit}
fit_prior <- praxis_bgm_fit(
  data = X_tgt,
  K = K,
  seed = 123L,
  prior_mus = prior_mus_src,
  prior_Sigmas = prior_Sigmas_src,
  prior_weights = prior_weights_src,
  beta = 1e-4,
  tol = 1e-4,
  max_iters = 300L,
  verbose = FALSE,
  prior_mus_variance = 1.5,
  num_samples = 64L,
  data_precision_int = 1L,
  likelihood_temp = 1.0,
  rho_prec = 0.05,
  rho_mu = 1.0,
  elbo_eval_freq = 5L,
  num_iters = 120L,
  batch_size = 64L,
  early_stop = TRUE,
  patience = 5L
)

ari_prior <- mclust::adjustedRandIndex(y_tgt, fit_prior$assignments_zero_based)
cat(sprintf("Target ARI with transferred priors: %.3f\n", ari_prior))
```

## Praxis-BGM without priors

```{r praxis-no-prior-fit}
fit_noprior <- praxis_bgm_fit(
  data = X_tgt,
  K = K,
  seed = 123L,
  prior_mus = NULL,
  prior_Sigmas = NULL,
  prior_weights = NULL,
  beta = 1e-3,
  tol = 1e-4,
  max_iters = 300L,
  verbose = FALSE,
  prior_mus_variance = 10.0,
  num_samples = 64L,
  data_precision_int = 1L,
  likelihood_temp = 1.0,
  rho_prec = 0.05,
  rho_mu = 1.0,
  elbo_eval_freq = 5L,
  num_iters = 120L,
  batch_size = 64L,
  early_stop = TRUE,
  patience = 5L
)

ari_noprior <- mclust::adjustedRandIndex(y_tgt, fit_noprior$assignments_zero_based)
cat(sprintf("Target ARI without priors: %.3f\n", ari_noprior))
```

## Summary

```{r summary}
cat("=== SUMMARY ===\n")
cat(sprintf("Praxis-BGM with transferred priors: %.3f\n", ari_prior))
cat(sprintf("Praxis-BGM without priors:         %.3f\n", ari_noprior))
cat(sprintf("Source-trained QDA baseline:       %.3f\n", ari_qda))
```

## Bayes-factor feature selection

```{r bf-selection}
bf_results <- praxis_bgm_bf_selection(
  model = fit_prior,
  data = X_tgt,
  top_n = 40L,
  visual = FALSE
)

top_features <- bf_results$top_features
bf_classification <- bf_results$classification
causal_features <- seq_len(C)
overlap_count <- length(intersect(top_features, causal_features))

top_features[1:10]
bf_classification
cat("Top-40 overlap with causal features:", overlap_count, "\n")
```

## Notes

```{r notes, echo=FALSE}
cat(
  paste(
    "- `fit_prior$assignments_zero_based` is used for exact comparison with the zero-based simulated labels.",
    "- `praxis_bgm_bf_selection()` returns one-based feature indices in `top_features` for R users.",
    "- The underlying BF scoring logic is still the Python Praxis-BGM implementation.",
    sep = "\n"
  )
)
```
