---
title: "Praxis-BGM R Wrapper"
author: "Qiran Jia"
date: "2026-04-20"
output:
  html_document: default
---

## Overview

This document describes the updated R interface for the current Python
`praxis_bgm` package. The repository now keeps the wrapper logic in a single
script:

- `R/praxis_bgm_interface.R`

That script is the source of truth for:

- `praxis_bgm_fit()`
- `praxis_bgm_bf_selection()`
- the simulation helpers used in tutorials and tests

The wrapper now matches the current Python constructor, including:

- `prior_weights` instead of `prior_pis`
- `init_pis` for initialization weights
- `freeze_A_zeros` for adjacency-mask freezing
- `data_precision_int`, `likelihood_temp`, `rho_prec`, `rho_mu`, and
  `elbo_eval_freq`

The BF-selection logic itself is unchanged from the Python package.

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)

library(reticulate)

praxis_python <- Sys.getenv("RETICULATE_PYTHON", unset = "")
if (nzchar(praxis_python)) {
  use_python(praxis_python, required = TRUE)
} else {
  use_condaenv("Praxis_env", required = TRUE)
}

source("R/praxis_bgm_interface.R")
```

```{r py-config}
py_config()
```

## Wrapper entry points

```{r signatures}
args(praxis_bgm_fit)
args(praxis_bgm_bf_selection)
```

## Minimal example

```{r example-data}
set.seed(123)

n_per_cluster <- 40L

cluster_1 <- cbind(
  rnorm(n_per_cluster, mean = -2.0, sd = 0.45),
  rnorm(n_per_cluster, mean = -1.8, sd = 0.50)
)

cluster_2 <- cbind(
  rnorm(n_per_cluster, mean = 2.1, sd = 0.50),
  rnorm(n_per_cluster, mean = 1.9, sd = 0.45)
)

X <- rbind(cluster_1, cluster_2)
colnames(X) <- c("feature_1", "feature_2")

prior_mus <- rbind(
  c(-1.9, -1.9),
  c(2.0, 2.0)
)

prior_Sigmas <- array(0, dim = c(2, 2, 2))
prior_Sigmas[1, , ] <- diag(c(1.0, 1.0))
prior_Sigmas[2, , ] <- diag(c(1.0, 1.0))
```

```{r example-fit}
fit <- praxis_bgm_fit(
  data = X,
  K = 2L,
  seed = 42L,
  prior_mus = prior_mus,
  prior_Sigmas = prior_Sigmas,
  prior_weights = c(0.5, 0.5),
  beta = 1e-3,
  tol = 1e-4,
  max_iters = 60L,
  num_iters = 8L,
  batch_size = 40L,
  num_samples = 8L,
  data_precision_int = 1L,
  likelihood_temp = 1.0,
  rho_prec = 0.08,
  rho_mu = 1.0,
  elbo_eval_freq = 1L,
  early_stop = TRUE,
  patience = 2L,
  verbose = FALSE
)
```

## Results

```{r results}
fit$assignments[1:10]
fit$assignments_zero_based[1:10]
fit$learned_weights
fit$model_summary
tail(fit$elbo_history, 3)
```

## Bayes-factor feature selection

```{r bf-selection}
bf <- praxis_bgm_bf_selection(
  model = fit,
  data = X,
  top_n = 2L,
  visual = FALSE
)

bf$top_features
bf$top_features_zero_based
bf$classification
```

## Notes

```{r notes, echo=FALSE}
cat(
  paste(
    "- `assignments` are returned as 1-based labels for R users.",
    "- `assignments_zero_based` preserves the raw Python labels.",
    "- `top_features` are returned as 1-based feature indices, and `top_features_zero_based` preserves the Python indices.",
    "- `model` remains a live Python object for advanced downstream use.",
    sep = "\n"
  )
)
```
