---
title: "Praxis_R_Turtorial"
author: "Qiran Jia"
date: "2025-10-01"
output:
  html_document: default
  pdf_document: default
---

## Overview

This tutorial shows how to source the `Praxis_R_Wrapper.R` script and reproduce the
Python tutorial workflow in R, including simulation, baselines, and Praxis-BGM fitting.

## Setup

```{r setup, include=FALSE}
source("Praxis_R_Wrapper.R")
library(reticulate)
library(MASS)
library(mclust)
library(proxy)
library(clue)

np <- import("numpy", delay_load = TRUE)
```

If these R packages are missing, install them first:

```{r install-packages, eval=FALSE}
install.packages(c("MASS", "mclust", "proxy", "clue"))
```

## Simulate SOURCE data

```{r helpers}
generate_overlapping_gmm_samples <- function(
  n_components,
  n_causal,
  n_features,
  n_samples,
  mean_shift = 0.4,
  random_seed = NULL
) {
  if (!is.null(random_seed)) {
    set.seed(random_seed)
  }

  base_mean <- rep(0, n_causal)
  causal_means <- vapply(
    seq_len(n_components),
    function(x) base_mean + rnorm(n_causal) * mean_shift,
    numeric(n_causal)
  )
  causal_means <- t(causal_means)

  samples_per_component <- n_samples %/% n_components
  samples_list <- vector("list", n_components)
  labels_list <- vector("list", n_components)

  for (k in seq_len(n_components)) {
    causal <- MASS::mvrnorm(samples_per_component, causal_means[k, ], diag(n_causal))
    non_causal <- matrix(
      rnorm(samples_per_component * (n_features - n_causal)),
      ncol = n_features - n_causal
    )
    samples_list[[k]] <- cbind(causal, non_causal)
    labels_list[[k]] <- rep(k - 1L, samples_per_component)
  }

  samples <- do.call(rbind, samples_list)
  labels <- unlist(labels_list)

  full_means <- matrix(0, nrow = n_components, ncol = n_features)
  full_means[, seq_len(n_causal)] <- causal_means

  list(samples = samples, labels = labels, full_means = full_means, covs = NULL)
}

randomly_shift_means <- function(true_means, shift_magnitude, percentage, random_seed) {
  set.seed(random_seed)
  shifted <- true_means
  n_clusters <- nrow(shifted)
  n_features <- ncol(shifted)
  n_shift <- as.integer(percentage * n_features)
  for (k in seq_len(n_clusters)) {
    idx <- sample(seq_len(n_features), size = n_shift, replace = FALSE)
    shifted[k, idx] <- shifted[k, idx] + rnorm(n_shift) * shift_magnitude
  }
  shifted
}

l2_norm_with_alignment <- function(est_means, true_means) {
  cost <- as.matrix(proxy::dist(est_means, true_means, method = "Euclidean"))
  assignment <- clue::solve_LSAP(cost)
  aligned_est <- est_means[seq_len(nrow(est_means)), , drop = FALSE]
  aligned_true <- true_means[assignment, , drop = FALSE]
  norm(aligned_est - aligned_true, type = "F")
}
```

```{r source-sim}
K <- 4
P <- 100
C <- 40
N_src <- 800

source_data <- generate_overlapping_gmm_samples(
  n_components = K,
  n_causal = C,
  n_features = P,
  n_samples = N_src,
  random_seed = 123
)

X_src <- source_data$samples
y_src <- source_data$labels
true_means_src <- source_data$full_means

cat("X_src shape:", dim(X_src), "\n")
cat("True means (source) shape:", dim(true_means_src), "\n")
```

## Learning priors from SOURCE

```{r source-priors}
mus_post_src <- do.call(
  rbind,
  lapply(seq_len(K), function(k) {
    Xk <- X_src[y_src == (k - 1L), , drop = FALSE]
    if (nrow(Xk) == 0) {
      stop(sprintf("No samples for cluster %d in source data!", k))
    }
    colMeans(Xk)
  })
)

pis_src_emp <- as.numeric(table(factor(y_src, levels = 0:(K - 1L)))) / length(y_src)

cat("Empirical SOURCE π:", round(pis_src_emp, 3), "\n")
cat("Empirical SOURCE means shape:", dim(mus_post_src), "\n")
```

## QDA baseline trained on SOURCE

```{r qda-source}
qda <- MASS::qda(X_src, grouping = as.factor(y_src))
cat("QDA fitted on SOURCE.\n")
```

## Simulate TARGET data (domain shift)

```{r target-sim}
N_tgt <- 200
mean_shift_tgt <- 0.3
shift_percentage <- 0.3

true_means_tgt <- randomly_shift_means(
  true_means_src,
  shift_magnitude = mean_shift_tgt,
  percentage = shift_percentage,
  random_seed = 999
)

set.seed(999)
samples_per_component_tgt <- N_tgt %/% K

X_tgt_list <- vector("list", K)
y_tgt_list <- vector("list", K)
for (k in seq_len(K)) {
  causal_tgt <- MASS::mvrnorm(
    samples_per_component_tgt,
    true_means_tgt[k, seq_len(C)],
    diag(C)
  )
  non_causal_tgt <- matrix(
    rnorm(samples_per_component_tgt * (P - C)),
    ncol = P - C
  )
  X_tgt_list[[k]] <- cbind(causal_tgt, non_causal_tgt)
  y_tgt_list[[k]] <- rep(k - 1L, samples_per_component_tgt)
}

X_tgt <- do.call(rbind, X_tgt_list)
y_tgt <- unlist(y_tgt_list)

cat("X_tgt shape:", dim(X_tgt), "\n")
cat("True means (target) shape:", dim(true_means_tgt), "\n")
```

## QDA baseline on TARGET

```{r qda-target}
qda_preds <- predict(qda, X_tgt)$class
ari_qda <- mclust::adjustedRandIndex(y_tgt, as.integer(as.character(qda_preds)))
cat(sprintf("[TARGET] ARI QDA (source-trained, supervised): %.3f\n", ari_qda))
```

## TARGET: Praxis-BGM with transferred priors

```{r praxis-priors}
result_prior <- praxis_bgm_fit(
  data = X_tgt,
  K = K,
  seed = 123,
  prior_mus = mus_post_src,
  prior_pis = NULL,
  beta = 1e-4,
  tol = 1e-4,
  max_iters = 300,
  verbose = TRUE,
  prior_mus_variance = 1.5,
  num_samples = 64,
  enforce_mask = FALSE,
  mask_space = "precision",
  spd_eps = 1e-6,
  num_iters = 120,
  batch_size = 64,
  early_stop = TRUE,
  patience = 5
)

pred_prior <- result_prior$model$predict(np$array(X_tgt, dtype = "float32"))
yhat_prior <- reticulate::py_to_r(pred_prior[[1]])
ari_prior <- mclust::adjustedRandIndex(y_tgt, yhat_prior)
cat(sprintf("[TARGET] ARI with transferred priors: %.3f\n", ari_prior))
```

## TARGET: Praxis-BGM without priors (baseline)

```{r praxis-noprior}
result_noprior <- praxis_bgm_fit(
  data = X_tgt,
  K = K,
  seed = 123,
  prior_mus = NULL,
  prior_Sigmas = NULL,
  beta = 1e-3,
  tol = 1e-4,
  max_iters = 300,
  verbose = TRUE,
  prior_mus_variance = 10.0,
  num_samples = 64,
  enforce_mask = FALSE,
  mask_space = "precision",
  spd_eps = 1e-6,
  num_iters = 120,
  batch_size = 64,
  early_stop = TRUE,
  patience = 5
)

pred_noprior <- result_noprior$model$predict(np$array(X_tgt, dtype = "float32"))
yhat_noprior <- reticulate::py_to_r(pred_noprior[[1]])
ari_noprior <- mclust::adjustedRandIndex(y_tgt, yhat_noprior)
cat(sprintf("[TARGET] ARI without priors: %.3f\n", ari_noprior))
```

## Final comparison

```{r summary}
cat("=== SUMMARY ===\n")
cat(sprintf("ARI NGVI (with transferred priors): %.3f\n", ari_prior))
cat(sprintf("ARI NGVI (no priors):             %.3f\n", ari_noprior))
cat(sprintf("[TARGET] ARI QDA (source-trained, supervised): %.3f\n", ari_qda))
```
